---
title: "Structural Drivers of Financial Nihilism"
subtitle: "A Cohort Analysis of Income, Homeownership, and Economic Attainability"
author: "Olivia Shipley"
date: "`r Sys.Date()`"
output: 
    rmdformats::downcute:
    self_contained: true
    code_folding: show
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ipumsr)
library(dplyr)
library(data.table)
```


```{r}
ddi <- read_ipums_ddi("~/Desktop/DA6833/FinancialNihilism/usa_00002.xml")
data <- read_ipums_micro(ddi)
```

```{r}
head(data)
nrow(data)
names(data)

table(data$YEAR)
```

```{r}
setDT(data)
data <- data[AGE >= 18]
```

```{r}
# Defining cohort age window and create birth year
data <- data %>%
  mutate(BIRTHYEAR = YEAR - AGE,
         GENERATION = case_when(
           BIRTHYEAR >= 1928 & BIRTHYEAR <= 1945 ~ "Silent",
           BIRTHYEAR >= 1946 & BIRTHYEAR <= 1964 ~ "Boomer",
           BIRTHYEAR >= 1965 & BIRTHYEAR <= 1980 ~ "GenX",
           BIRTHYEAR >= 1981 & BIRTHYEAR <= 1996 ~ "Millennial",
           BIRTHYEAR >= 1997 & BIRTHYEAR <= 2012 ~ "GenZ",
           TRUE ~ "Other"
         ))

# Check the generation distribution
table(data$GENERATION)
```

```{r}
data <- data[GENERATION != "Other"]
table(data$GENERATION)
```

```{r}
homeown_by_gen <- data[, .(
  homeownership_rate = mean(OWNERSHP == 1, na.rm = TRUE)
), by = GENERATION]

homeown_by_gen[order(homeownership_rate)]
```
#not seperating cohorts by generations here^ this table is collapsing across all years, so it's mixing age effects with cohort effects. 


#homeownership rate by generation and year, so you can see the trend over time
```{r}
homeown_by_gen_year <- data[, .(
  homeownership_rate = mean(OWNERSHP == 1, na.rm = TRUE)
), by = .(GENERATION, YEAR)]

homeown_by_gen_year[order(GENERATION, YEAR)]
```


```{r}
library(ggplot2)

ggplot(homeown_by_gen_year, aes(x = YEAR, y = homeownership_rate, color = GENERATION)) +
  geom_line() +
  geom_point() +
  labs(title = "Homeownership Rate by Generation Over Time",
       x = "Year", y = "Homeownership Rate", color = "Generation") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```
#plot is tracking homeownership through time, which is not exactly what we are looking for but there are some interesting obervations. 

#Silent Generation declining after ~2010 makes sense — they're dying or moving into care facilities, so homeownership drops at the tail.
Millennials dipping around 2015 before rising likely reflects the post-2008 housing crash fallout hitting them right as they entered prime buying years.
Gen Z's trajectory is very short — they only appear in 2019 onward, and the 2024 uptick may just reflect the older end of Gen Z aging into homeownership.





##25-35 Age Filter & Inflation Adjustment
```{r}
dt_prime <- data[AGE >= 25 & AGE <= 35]
```

```{r}
cpi_lookup <- data.table(
  YEAR = c(1970, 1980, 1990, 2000, 2010, 2015, 2019, 2022, 2024),
  cpi  = c(38.8, 82.4, 130.7, 172.2, 218.1, 237.0, 255.7, 292.7, 313.5)
)
```

```{r}
dt_prime[VALUEH == 9999999, valueh_real := NA]
dt_prime[HHINCOME == 9999999, hhinc_real := NA]
dt_prime[INCTOT == 9999999, inc_real := NA]
```

```{r}
dt_prime[inc_real < 0, .N]
dt_prime[, .N, by = OWNERSHP]
dt_prime[, .N, by = .(OWNERSHP == 0, GENERATION)]
```


#very small fraction of the dataset claimed a negative income (4k)
# the 0s likley represent people living in group quarters, prisions, co-ops, etc

# there is a clear generational gradient — Gen Z and Millennials are disproportionately in the "not in universe" category compared to older cohorts.

#However, claude said we should still exclude them from the main analysis, because we genuinely don't have valid housing status data for them. 

#younger cohorts are more likely to fall outside the own/rent binary, which may mean the analysis slightly understates the homeownership gap for Millennials and Gen Z.


```{r}
dt_prime <- dt_prime[OWNERSHP %in% c(1, 2)]
```

#weights !
```{r}
homeown_weighted <- dt_prime[, .(
  homeownership_rate = weighted.mean(OWNERSHP == 1, w = PERWT, na.rm = TRUE)
), by = GENERATION]

homeown_weighted[order(homeownership_rate)]
```

```{r}
dt_prime[, .(mean_perwt = mean(PERWT)), by = YEAR][order(YEAR)]
```
The U.S. population has grown significantly over time, so there are simply more 25–35 year olds alive today than there were in 1970-1990 when Boomers were in that age range.
So when you have a 1970 observation of a 25-year-old Boomer with PERWT = 100, that person represents 100 people from a much smaller total pool of 25–35 year olds. A 2022 observation of a 25-year-old Millennial with PERWT = 116 represents 116 people from a much larger total pool.
When you compute a weighted mean across all generations pooled together, the Millennial and Gen Z observations just vastly outnumber the Boomer and Silent ones in terms of total weighted population — not because of anything wrong with the data, but because there genuinely are more young people in the U.S. today than there were 50 years ago.
This is why the compression happens — older generations get numerically drowned out in a pooled weighted analysis even though their individual weights are similar.
And this is actually a good argument for why your primary results should present homeownership rates by generation separately, rather than in a single pooled model. The life-stage comparison (everyone at 25–35) is the right framing, and each generation should be interpreted on its own terms within that window.

Price-to-Income Ratio = VALUEH / INCTOT
A ratio of 3 used to be the rule of thumb — a house should cost no more than 3 years of gross income. As that ratio rises, affordability deteriorates.


#need to decide weither or not to use individual income or household income 
```{r}
dt_prime[inc_real > 0, price_to_income_ind := valueh_real / inc_real]
dt_prime[hhinc_real > 0, price_to_income_hh := valueh_real / hhinc_real]

summary(dt_prime[OWNERSHP == 1, .(price_to_income_ind, price_to_income_hh)])
```

#Historical Note: 
In 1970, single women owning homes was relatively rare — both culturally and legally. The Equal Credit Opportunity Act wasn't passed until 1974, meaning before that lenders could legally deny women credit without a male co-signer. So a single woman owning a home in 1970 was genuinely uncommon.

This creates an interesting asymmetry in your data:
-In 1970, a single male homeowner's individual income and household income are essentially the same thing — he's the sole earner and sole owner
-A married homeowner in 1970 was almost certainly a male primary earner with a spouse who may have had little or no individual income
-Single female homeowners in 1970 are rare enough that they may represent a select, unusually financially independent group

For single-person households, individual and household income are identical anyway — INCTOT and HHINCOME would be the same person. So the household income measure actually handles single homeowners just fine. The problem only arises for multi-person households where individual income misrepresents the household's actual purchasing power.
This is also a good argument for keeping marital status (MARST) as a control variable in your regression — it helps account for the fact that the relationship between income and homeownership looked very different for single vs. married individuals, and that dynamic has shifted significantly across your generational cohorts.

##DECISION:
Using price_to_income_hh as the affordability component of the FAI due to the skewed data from the single income homeowners being absurd^ 

##DECISION: 
A notable limitation of this analysis is its inability to directly capture student debt burden, which represents one of the most significant financial constraints facing Millennial and Gen Z households. While sources such as the Federal Reserve's Survey of Consumer Finances contain individual-level student debt data, they cannot be meaningfully merged with the IPUMS microdata used here at the individual level. Aggregate student debt statistics, available at the national level by year, were considered but ultimately excluded as they offer no individual-level variation and would function essentially as a year fixed effect in the regression framework. As a result, the debt burden component of this analysis is necessarily limited to mortgage-related costs among homeowners. This omission likely means that the financial constraints facing younger cohorts are understated in the models, and future research with richer debt data would be well positioned to extend these findings.




